The Economics of Wildfire Aviation

### 2023-09-03 manuscript/manuscript.qmd
added discussion content and started reorganizing

---
title: "The Economics of Wildfire Aviation"
format: 
  docx:
    reference-doc: custom-reference-doc.docx
    toc: true
    number-sections: true
bibliography: references.bib
csl: journal-of-agricultural-and-resource-economics.csl
---

@@

<!-- high level connecting wildfire to climate change; mitigation is not enough -->
<!-- consider new opening jumping straight into aviation or starting with overall cost -->

Changing climate is driving longer and more intense fire seasons. There is an urgent need to reduce wildfire risk in fire-prone regions of the US. Yet, the primary method for reducing risk at the landscape level involves the costly removal of combustible fuels. Despite significant progress, the number of acres in need of treatment is very large and not expected to be treated over the next decade [@thompson2023]. Wildfire suppression and response currently does and will continue to play an important role in protecting people, property, and other valued assets from wildfire.

@@

Over the past 10 years, the federal government spent an average of \$2.3 billion per year on wildfire suppression [@suppress]. Approximately 21% (\$493 million) of that is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available. Given the high cost of aviation, it is critical to determine whether the benefits exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel.

@@

The objective of this paper is to describe/investigate the economic aspects of aviation used for wildfire response. First, we will discuss the lack of objective measurement of the benefits of aviation that is necessary to subject their use to benefit cost analysis. Second, we will apply economic principles/theory to the procurement of aviation resources. While this paper will not analyze data or come to clear conclusions about the use of aviation, it will frame the economic problems and provide a path forward for how to solve them.

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CNW). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts specify that an aircraft is available for a short period of time (e.g., 3 days) at a daily cost higher than the EXU availability rate. The question is: how many of each contract should the USFS sign at the start of each fire season? The answer depends on expected use patterns of the aviation resources.

@@

Figure 2 illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.^[The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in figure 2. The trends in figure 2 suggest that given current usage patterns, the USFS could delay activating some LATs and VLATs until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

@@

>>>>

---
title: "The Economics of Wildfire Aviation"
format: 
  docx:
    reference-doc: custom-reference-doc.docx
    toc: false
    number-sections: true
bibliography: references.bib
csl: journal-of-agricultural-and-resource-economics.csl
---

@@

Over the past 10 years, the federal government has spent an average of \$2.3 billion per year on wildfire suppression [@suppress]. Approximately 21% (\$493 million) of that is spent on aviation resources [@stonesifer2021]. These costs involve hourly flight costs to operate the aircraft, the cost of retardant, and the cost to keep aircraft available and maintain it. Given the high cost of aviation, it is critical to determine whether the benefits of their application exceed the costs. Otherwise, resources could be better invested in other firefighting resources such as wildland firefighting engines, dozers, and personnel.

For the past decade, federal firefighting agencies have contracted with private companies that own and maintain firefighting aircraft. The structure of these contracts and the procurement mechanism influence how aviation resources are used and the costs borne by firefighting agencies.

The objective of this paper is to describe and analyze two economic aspects of aviation used for wildfire response. First, we discuss the lack of objective measurement of the benefits of aviation that is necessary to subject their use to benefit cost analysis. Second, we apply economic principles to the procurement of aviation resources and discuss how and where economic efficiency can be gained. While this paper will not analyze data or come to clear conclusions about the use of aviation, it will frame the economic problems and attempt to provide a path forward for how to solve them.

@@

@@

@@

The USFS determined that owning and maintaining its own fleet of aircraft is not worth the cost. In xxxx, they changed to a procurement model in which private firefighting aircraft are contracted for a period of time at a specified cost. The contacts specify a number of safety requirements, but also a daily availability price and an hourly flight price (cite Stoneseifer). The two forms of contract are Exclusive Use (EXU) and Call-When-Needed (CNW). Exclusive use contracts mandate that an aircraft is exclusively available to the contracting entity (e.g., US Forest Service) for a specified number of days at a daily cost. Call-when-needed contracts are more flexible and can be activated on short notice. However, CWN contacts generally specify a daily availability rate and hourly flight rate that is higher than EXU contracts. The question is: how many of each contract should the USFS sign at the start of each fire season? The answer depends on expected use patterns of the aviation resources.

@@

Figure 2 illustrates the variability and uncertainty of initial attack fires (bottom panel) and large fires being actively suppressed (extended attack - top panel) throughout the season. According to the 2022 USFS Schedule of Items, most large airtankers under exclusive use contracts started between March and May and were scheduled to end in September, October, or November.[^1] While the 18 large and very large airtankers under contract in 2022 span the dates when fire activity is expected to be high, they are not activated in a pattern that follows the number of IA and EA fires in figure 2. The trends in figure 2 suggest that given current usage patterns, the USFS could delay activating some LATs and VLATs until later in June. Call When Needed contracts could be used to handle surge capacity and unexpected increases in fire activity.

[^1]: The only publicly accessible document we could locate was from June 24, 2022 https://www.fs.usda.gov/Internet/FSE_DOCUMENTS/fseprd1041508.pdf

@@

# Discussion

<!-- high level connecting wildfire to climate change; mitigation is not enough -->

A changing climate is driving longer and more intense fire seasons. There is an urgent need to reduce wildfire risk in fire-prone regions of the US. Yet, the primary method for reducing risk at the landscape level involves the costly removal of combustible fuels. Despite significant progress, the number of acres in need of treatment is very large and not expected to be treated over the next decade [@thompson2023]. Wildfire suppression and response currently does and will continue to play an important role in protecting people, property, and other valued assets from wildfire. As federal firefighting agencies seek to improve efficient use of resources, the procurement and use of aviation resources presents several opportunities.

# Maybe material

-   more flexible CWN contracts: Current records suggest that CWN contracts are executed for 90 days. The federal firefighting agencies could benefit from shorter and more flexible contracts. If there were no transactions cost[^2], the ideal arrangement would allow the firefighting agency to use (and pay for) the aircraft only when they need it.

[^2]: By transactions cost, we mean the time and effort to arrange, verify, and execute the contract, the moving of aircraft, and any other logistical issues involved in using the aircraft.

For example, the default CWN contract could be one week with a requirement that the federal agency renew or release the aircraft before

-   sunk cost fallacy: The structure of the procurement model may influence how aviation resources are used. The daily availability price can range from \$30k to \$50k per day while the hourly flight cost ranges from \$xxk to \$xxk. Those involved in allocating aviation resources may view unused aircraft as wasteful since the daily availability rate is paid regardless of use.

-   The schedule of items suggests that there are relatively few private entities that own aircraft. Procurement auctions tend to be more economically efficient when there are more bidders competing for the limited contract. Auction theory suggests that the competition between the bidders over a limited number of contracts will incentivize them to reveal their true willingness to accept, which should equal their marginal cost of provision. Moreover, the auction efficiency assumes that each bidder's cost of provision is unknown to the other bidders. When the number of bidders is small and there is ample opportunity for the bidders to communicate in advance of submitting bids, it is more likely that the bidders will bid above their marginal cost of provision. Moreover, the aircraft owners repeatedly participate in this annual procurement process allowing them to learn about their competitors and the government's willingness to pay. In principle, these forces may lead to higher procurement costs. However, these auction inefficiencies may be unavoidable since the ownership and maintenance of large aircraft is capital intensive, which serves as a barrier to new entrants in the market.[^3]

[^3]: The McDonnell Douglas DC 10, one of the largest active airtankers, is currently listed for sale for \$2.75 million (https://myemail.constantcontact.com/McDonnell-Douglas-DC-10-30-Freighter-For-Sale--Price-Reduction-.html?soid=1102673403687&aid=zTC4sV-oTNI, accessed September 9, 2023).

Moreover, the federal agencies are not the only customers of companies that own firefighting aircraft. Indeed, these companies increasingly sell their services to other countries (e.g., Canada and Australia) as well as US state agencies. For example, the Colorado Division of Fire Prevention and Control has signed exclusive use contracts for a BA-146 airtanker provided by Neptune Aviation [@gabbert2021].
